Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-08-18 paper/ifo-complexAudio.tex
some editing in wandering tone and complex audio paper tex

% Optical microphone
% aim of optical microphone: play back injected sounds
A natural successor to simple tomes is complex audio, such as music and speech. 
Complex audio signals are not quasi-monochromatic, unlike continuous gravitational wave signals, so the Viterbi algorithm as implemented in Section~\ref{sec:viterbi_wandering} is not applicable directly.
Instead, in this section, we explore how to use the interferometer as an `optical microphone' to detect and play back complex audio signals, using light to capture sound. 
In doing so, we experiment with a hierarchy of passive filters which suppress noise yet make no assumption about the specific form of the signal, unlike the Fourier-based maximum likelihood matched filter tuned to sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood}. 
The device has precedence in the laser microphones~\cite{laser_microphone} used in the defence industry which operate on a variety of related principles.
It is also analogous to the detection of unmodelled gravitational-wave signals, e.g. broadband bursts from supernovae~\cite{}\han{to do - add citation}. 
%It is also analogous to the detection of unmodelled gravitational wave signals, e.g. broadband bursts from supernovae. [refs; my last sentence can be improved]

@@

We find that the raw output of the optical microphone, collected without any further processing (except anti-aliasing), is noisy, with a loud bass hum throughout.
Each $10\,{\rm s}$ Fourier spectrum shows dominant AC mains noise with power ranging from the fundamental of the $50\,{\rm Hz}$ mains power up to and beyond its $8$th harmonic. \jam{[non-linear appendix, Andrew says to cite Feynman lectures on non-linear oscillations]}
This mains power signal is present in the power spectral density of background noise as shown in the top panel of Figure~\ref{fig:psd_noise}.
The mains power signal is far weaker but still present in the background noise taken with the photodiode in darkness, suggesting that ambient lighting has a large contribution.
The origin of the mains power signal is likely to be from the photodiode signal, or environmental noise from the rooms lighting and air conditioning, or a combination. 
Outside the mains noise, the power appears to peak broadly around 750 Hz, the origin or significance of which we don't know.

@@

The next step is to apply filters to the data. 
The ideal filter would be one that 
\begin{itemize}
\item[i)] completely attenuates the undesired parts of the spectrum, 
\item[ii)] does not change the rest of the spectrum, and 
\item[iii)] has smooth edges so as to not damage the signal after convolution. 
\end{itemize}
These three conditions are contradictory and any filter must compromise between them.

@@

For a low-pass Butterworth filter with cut-off frequency $\omega_c$ and gain $\varepsilon$, the frequency response (attenuation at each frequency) as a function of frequency $\omega$ is:
\begin{equation}
\label{eq:butterworth}
H(\omega) = \left(1+\varepsilon^2 \left( \frac{\omega}{\omega_c} \right)^{2n}\right)^{-\frac{1}{2}}
\end{equation}

\jam{[Andrew wanted a quantitative comment about what the Butterworth does to the PSD, what do I use?]}
To construct a Butterworth band-pass filter, the low-pass filter in Eq.~\ref{eq:butterworth} is combined with a high-pass filter of similar form. 
The best performing band-pass filter we find is a fifth order ($n = 5$) Butterworth filter over the range $150\,{\rm Hz}$--$3\,{\rm kHz}$ (beyond $2\,{\rm kHz}$ speech information content drops significantly~\cite{speech_intelligibility}). 
The response of the filter used is shown in Figure~\ref{fig:butterworth} and its effect on the power spectral density of background noise is shown in the bottom panel of Figure~\ref{fig:psd_noise}.
Band-passing the signal does not address the remaining harmonics above $100\,{\rm Hz}$. 
However, it does suppress other noise at higher frequencies. 
The Butterworth filter performs well at maintaining source intelligibility as it is optimised to be `maximally flat' within the band region (prioritising condition (ii) above).
However recordings of speech still have low intelligibility (the speech can't be understood) as the filter does not improve beyond the source.
% They can be recognised as speech without being understood.
Looking forward to Sec.~\ref{sec:logmmse}, results for the Butterworth filter are shown in combination with a speech enhancement technique to address this.

@@

We also consider a combined notch filter and Wiener filter.
Timeseries and frequency spectrum results are shown in Figs.~\ref{fig:timeNotchWiener} and ~\ref{fig:freqNotchWiener} respectively. 
The original signal is identical to that shown in Fig.~\ref{fig:noiseSubtract}.
The first and second panels of Fig.~\ref{fig:timeNotchWiener} show the original audio signal and the recording after passing through the optical microphone (i.e. the timeseries of the data shown in the first two panels of Fig.~\ref{fig:noiseSubtract}).

>>>>

% Optical microphone
% aim of optical microphone: play back injected sounds
A natural successor to single and wandering tones is complex audio, such as music and speech. 
Complex audio signals are not quasi-monochromatic, unlike continuous gravitational wave signals, so the Viterbi algorithm as implemented in Section~\ref{sec:viterbi_wandering} is not applicable directly.
Instead, in this section, we explore how to use the interferometer as an `optical microphone' to detect and play back complex audio signals, using light to capture sound. 
In doing so, we experiment with a hierarchy of passive filters which suppress noise yet make no assumption about the specific form of the signal, unlike the Fourier-based maximum likelihood matched filter tuned to sinusoidal signals in Section~\ref{sec:single_tone} and Appendix~\ref{app:sinusoid_likelihood}. 
The device has precedence in the laser microphones~\cite{laser_microphone} used in the defence industry which operate on a variety of related principles.
%It is also analogous to the detection of unmodelled gravitational-wave signals, e.g. broadband bursts from supernovae~\cite{}\han{to do - add citation}. 
%It is also analogous to the detection of unmodelled gravitational wave signals, e.g. broadband bursts from supernovae. [refs; my last sentence can be improved]

@@

We find that the raw output of the optical microphone, collected without any further processing (except anti-aliasing), is noisy, with a loud bass hum throughout.
Each $10\,{\rm s}$ Fourier spectrum shows dominant AC mains noise with power ranging from the fundamental of the $50\,{\rm Hz}$ mains power up to and beyond its $8$th harmonic, likely due to the non-linearity discussed in Appendix~\ref{app:intensity_derivation} (see also Ref.~\cite{feynman}). 
This mains power signal is present in the power spectral density of background noise as shown in the top panel of Figure~\ref{fig:psd_noise}.
The mains power signal is far weaker but still present in the background noise taken with the photodiode in darkness, suggesting that ambient lighting has a large contribution.
The origin of the mains power signal is likely to be from the photodiode signal, or environmental noise from the rooms lighting and air conditioning, or a combination. 
Outside the mains noise, the power appears to peak broadly around 750 Hz, the origin or significance of which we don't know.

@@

The next step is to apply filters to the data. 
The ideal filter would be one that:
%\begin{itemize}
i) completely attenuates the undesired parts of the spectrum, 
ii) does not change the rest of the spectrum, and 
iii) has smooth edges so as to not damage the signal after convolution. 
%\end{itemize}
These three conditions are contradictory and any filter must compromise between them.

@@

We also apply a Butterworth filter. 
A low-pass Butterworth filter with cut-off frequency $f_c$ and gain $\varepsilon$, has a frequency response (attenuation at each frequency)
\begin{equation}
\label{eq:butterworth}
H(\omega) = \left(1+\varepsilon^2 \left( \frac{f}{f_c} \right)^{2n}\right)^{-\frac{1}{2}}
\end{equation}
% commenting to send to co-authors
%\jam{[Andrew wanted a quantitative comment about what the Butterworth does to the PSD, what do I use?]}
To construct a Butterworth band-pass filter, the low-pass filter in Eqn.~\ref{eq:butterworth} is combined with a high-pass filter of similar form. 
The best performing band-pass filter we find is a fifth order ($n = 5$) Butterworth filter over the range $150\,{\rm Hz}$--$3\,{\rm kHz}$ (beyond $2\,{\rm kHz}$ speech information content drops significantly~\cite{speech_intelligibility}). 
The effect of the Butterworth band-pass filter on the power spectral density of the background noise is shown in the bottom panel of Fig.~\ref{fig:psd_noise}, where we see that it reduces the amplitude of the mains harmonics below $150\,{\rm Hz}$ 
The response of the filter is shown in Fig.~\ref{fig:butterworth}.
Band-passing the signal does not address the remaining harmonics above $100\,{\rm Hz}$. 
However, it does suppress other noise at higher frequencies. 
The Butterworth filter performs well at maintaining source intelligibility as it is optimised to be `maximally flat' within the band region (prioritising condition ii) above).
However recordings of speech still have low intelligibility (the speech can't be understood).
Looking forward to Sec.~\ref{sec:logmmse}, results for the Butterworth filter are shown in combination with a speech enhancement technique to address this.

@@

We also consider a combined notch filter and Wiener filter. 
Here we describe our use of these methods with timeseries and frequency spectrum results shown in Figs.~\ref{fig:timeNotchWiener} and ~\ref{fig:freqNotchWiener} respectively. 
The original signal is identical to that shown in Fig.~\ref{fig:noiseSubtract} where the first and second panels of Fig.~\ref{fig:timeNotchWiener} show the original audio signal and the recording after passing through the optical microphone respectively (i.e. the timeseries of the data shown in the first two panels of Fig.~\ref{fig:noiseSubtract}).

### 2020-08-18 paper/ifo-wanderingTone.tex
some editing in wandering tone and complex audio paper tex

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each box represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The colour of each circle indicates the likelihood that a signal is present in the box, where a lighter colour corresponds to higher likelihood. 
AT each time step, the paths can move up one $f$ step, move down one $f$ step, or stay at the same $f$ step (shown by lines in the diagram). 
Each circles has three paths leading to it indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by equation~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

\item Moving to $t_2$, the path with the highest value of the recursive relation in equation~\ref{eqn:recursiveViterbi} is selected for each $f_j$ as shown by the solid lines from $t_1$ to $t_2$ in Fig.~\ref{fig:viterbi}.

@@

\item At $t=t_{N_t}$, each $f_j$ has a value ${\rm Pr}[f(t_{N_t})=f_j]$ (equation~\ref{eqn:recursiveViterbi}). 
The algorithm selects the terminating frequency $f(t_{N_t})$ with the highest value, marked (b) in Fig.~\ref{fig:viterbi}.

\item The Viterbi path is then found by backtracking along the best paths at each time step.
\jam{[Andrew mentioned providing a closed form here, which I'm not sure exists. The code works by storing three arrays: the weights, the score for the best path to a given vertex, and the index of the previous vertex along that path. Backtracking comprises of looping through that third array from the vertex in the last row with the highest score from the second array. I don't think we need to explain it any further than that.]}
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighted in orange in fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively maximized over all terminating frequency bins. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
We also consider the root mean square (RMS) fractional error along the path. Given $i=0,1,...,N_f$ columns of the injected frequency $I_i$ and recovered (Viterbi) frequency $R_i$, the RMS fractional error is given by:
\begin{equation}
\sqrt{ \frac{1}{N_f+1} \sum_{i=0}^{N_f} \frac{(I_i - R_i)^2}{{I_i}^2} }
\end{equation}
Computing this along the Viterbi path we find an error of 0.082, which indicates fair but not total recovery of the injected signal.

>>>>

\begin{figure}
\includegraphics[width=0.49\textwidth]{figures/viterbiDiagram.pdf}
\caption{\label{fig:viterbi}
A schematic diagram of the Viterbi algorithm. 
Each box represents an element in the time-frequency grid starting from the instant $t_0$ on the left and ending at $t_{N_t}$ on the right. 
Vertically, the grid starts from frequency $f_0$ at the bottom and ending at $f_{N_f}$ at the top. 
The colour of each circle indicates the likelihood that a signal is present in the box, where a lighter colour corresponds to higher likelihood. 
AT each time step, the paths can move up one $f$ step, move down one $f$ step, or stay at the same $f$ step (shown by lines in the diagram). 
Each circles has three paths leading to it indicated by the lines. 
The best path to each circle at each $t_i$ is highlighted in black. 
Some routes through the grid are found to be dead ends, such as the path ending at $t_1$ marked (a). 
At $t=t_{N_t}$ the algorithm chooses the terminating frequency circle which has the highest value given by Eqn.~\ref{eqn:recursiveViterbi}, marked (b). 
The Viterbi path is the path leading to this circle, highlighted in orange from (b) to the start at (c). 
}
\end{figure}

@@

\item Moving to $t_2$, the path with the highest value of the recursive relation in Eqn.~\ref{eqn:recursiveViterbi} is selected for each $f_j$ as shown by the solid lines from $t_1$ to $t_2$ in Fig.~\ref{fig:viterbi}. 
The origins are stored for each grid point.

@@

\item At $t=t_{N_t}$, each $f_j$ has a value ${\rm Pr}[f(t_{N_t})=f_j]$ (Eqn.~\ref{eqn:recursiveViterbi}). 
The algorithm selects the terminating frequency $f(t_{N_t})$ with the highest value, marked (b) in Fig.~\ref{fig:viterbi}.

\item The Viterbi path is then found by backtracking along the stored best paths at each time step (see also Appendix~\ref{app:viterbi}). 
% come back to this. 
%\jam{[Andrew mentioned providing a closed form here, which I'm not sure exists. The code works by storing three arrays: the weights, the score for the best path to a given vertex, and the index of the previous vertex along that path. Backtracking comprises of looping through that third array from the vertex in the last row with the highest score from the second array. I don't think we need to explain it any further than that.]}
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighted in orange in fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot and white-cross markers show the injected signal and recovered Viterbi path respectively maximized over all terminating frequency bins. 
The Viterbi path provides a good match to the injected frequency. 
%To the eye, the Viterbi algorithm easily recovers the injected frequency over time due to the high signal-to-noise ratio (SNR) of this experiment \han{What was the SNR? / can we quantify this?}.
Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
We also consider the root mean square (RMS) fractional error $E_{\rm rms}$ along the path. Given $i=0,1,...,N_f$ columns of the injected frequency $I_i$ and recovered (Viterbi) frequency $R_i$, $E_{\rm rms}$ is given by
\begin{equation}
E_{\rm rms} = \sqrt{ \frac{1}{N_f+1} \sum_{i=0}^{N_f} \frac{(I_i - R_i)^2}{{I_i}^2} }
\end{equation}
Computing this along the Viterbi path we find $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.
